---
title: "Cross-language reproducibility"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cross-language reproducibility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(pilotr)
```

## The challenge

A toolkit that spans R and Python faces a subtle obstacle. The two ecosystems use different
default random-number generators (R's Mersenne-Twister with inversion, NumPy's PCG64), so a
direct port produces different data from the same seed. That would undermine the guarantee
that a design authored once can be reproduced anywhere.

## A shared generator

pilotr sets aside the native RNGs and provides a shared generator implemented identically
in both languages. Uniform deviates come from L'Ecuyer's (1988) combined linear congruential
generator, whose arithmetic stays below $2^{53}$ and is therefore exact in IEEE-754 doubles
(R) and Python integers alike. Normal deviates use Wichura's (1988) Algorithm AS 241
inverse-CDF, the algorithm R's `qnorm()` uses, so deviates agree to full double precision.
The remaining draws (Cholesky-correlated random effects, inverse-CDF Poisson and ordinal
draws, Marsaglia–Tsang gamma draws for the Beta family, a Fisher–Yates item sampler for
partial crossing) are derived from those two through a documented, identical consumption
order.

As a result, given the same specification and seed, the R and Python implementations
produce bit-identical data sets. The specification, rather than the implementation, serves
as the source of truth.

## Determinism in R

The same specification and seed always give the same data.

```{r}
spec <- build_spec(list(
  name = "demo", seed = 2024, design_kind = "between", n_subject = 200,
  factor_name = "group", lev1 = "a", lev2 = "b",
  intercept = 100, effect = 5, family = "gaussian", resp_name = "score", sigma = 10))

isTRUE(all.equal(simulate_design(spec), simulate_design(spec)))
```

Changing the seed changes the draws, while the structure and the ground-truth parameters stay
fixed.

```{r}
spec2 <- spec; spec2$seed <- spec$seed + 1L
identical(simulate_design(spec)$score, simulate_design(spec2)$score)
```

## The same design in Python

The Python package reads the identical specification and produces the identical data. The
code below is not evaluated here. It shows the Python you would run, and its output matches
the R data to the last bit.

```python
from pilotr import simulate, load_spec

# the same design.json the R package (and the app) use
data = simulate("design.json")
```

The repository includes a parity check that simulates the worked example designs in both
languages and reports the maximum absolute difference per design.

```{r, echo = FALSE}
# Output of python/examples/parity_check.py --csv, precomputed and shipped with the
# package: the comparison needs both a Python and an R installation, so it cannot run
# at vignette build time.
# max_abs_diff is read as text: as a number, kable would round 4.885e-15 to 0 and the
# table would claim a stricter result than the check actually produced.
knitr::kable(read.csv("parity-summary.csv", colClasses = c(max_abs_diff = "character")),
             format = "html", table.attr = 'class="table data-output"')
```

The difference is exactly zero for every design whose responses are rounded, and about
5e-15 for the unrounded continuous design, where R's CSV writer prints 15 significant
digits. That residual is a number-formatting artefact of the comparison rather than a
divergence between the generators, and it is what the check's 1e-6 tolerance is there to
absorb. There are no categorical mismatches anywhere. The uniform stream is exact by
construction, since every intermediate product stays below $2^{53}$, and the full pipeline has
been observed bit-identical across the platforms tested. The parity check can be rerun on any
new platform, so a design verified on a laptop can be confirmed on an HPC node before a large
run.

## Why this matters

Because the specification is the source of truth and the generator is shared, a design can be
authored once (in the app, in R or in Python) and reproduced everywhere. It can be preregistered
as a small JSON file that anyone can run, scaled from a laptop to a cluster without drift, and
handed between collaborators who work in different languages.

This property turns an informal request to simulate some data into a reproducible, shareable
methodological artefact.
