---
title: "Cross-language reproducibility"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cross-language reproducibility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "")
# Console colour carries no meaning on a rendered page. pkgdown turns it on for
# its own build, and the escape sequences then reach the reader as literal text,
# so colour is switched off here for a plain vignette render and a site build
# alike. The fixed width keeps printed output inside the documentation column.
options(cli.num_colors = 1, cli.hyperlink = FALSE, crayon.enabled = FALSE,
        width = 80)
```

```{r setup}
library(pilotr)
```

## The challenge

A toolkit that spans R and Python faces a subtle obstacle. The two ecosystems use different
default random-number generators (R's Mersenne-Twister with inversion, NumPy's PCG64), so a
direct port produces different data from the same seed. That would undermine the guarantee
that a design authored once can be reproduced anywhere.

## A shared generator

pilotr sets aside the native RNGs and provides a shared generator implemented identically
in both languages. Uniform deviates come from L'Ecuyer's (1988) combined linear congruential
generator, whose arithmetic stays below $2^{53}$ and is therefore exact in IEEE-754 doubles
(R) and Python integers alike. Normal deviates use Wichura's (1988) Algorithm AS 241
inverse-CDF, the algorithm R's `qnorm()` uses, so deviates agree to full double precision.
The remaining draws (Cholesky-correlated random effects, inverse-CDF Poisson and ordinal
draws, Marsaglia–Tsang gamma draws for the Beta family, a Fisher–Yates item sampler for
partial crossing) are derived from those two through a documented, identical consumption
order.

As a result, given the same specification and seed, the R and Python implementations
produce identical data sets: bit for bit for rounded and Gaussian responses, and within a few
units in the last place where an unrounded family applies `exp()` or `log()`, as the parity
check below quantifies. The specification, rather than the implementation, serves
as the source of truth.

## Determinism in R

The same specification and seed always give the same data.

```{r}
spec <- build_spec(list(
  name = "demo", seed = 2024, design_kind = "between", n_subject = 200,
  factor_name = "group", lev1 = "a", lev2 = "b",
  intercept = 100, effect = 5, family = "gaussian",
  resp_name = "score", sigma = 10))

isTRUE(all.equal(simulate_design(spec), simulate_design(spec)))
```

Changing the seed changes the draws, while the structure and the ground-truth parameters stay
fixed.

```{r}
spec2 <- spec; spec2$seed <- spec$seed + 1L
identical(simulate_design(spec)$score, simulate_design(spec2)$score)
```

## The same design in Python

The Python package reads the identical specification and produces the identical data. The
code below is not evaluated here. It shows the Python you would run, and its output matches
the R data to the last bit.

```python
from pilotr import simulate, load_spec

# the same design.json the R package (and the app) use
data = simulate("design.json")
```

The repository includes a parity check that simulates the worked example designs in both
languages and reports the maximum absolute difference per design. This is the CSV-level check,
read at the precision a written file carries and with a tolerance of 1e-6. The stricter one is
the harness under `tools/parity/`, which dumps every value at 17 significant digits, accounts
for each differing cell in units in the last place against a recorded allowance, and runs as a
gate in continuous integration. The table below is the readable summary; that harness is what
the bit-identical claim actually rests on.

```{r, echo = FALSE}
# Output of python/examples/parity_check.py --csv, precomputed and shipped with
# the package: the comparison needs both a Python and an R installation, so it
# cannot run at vignette build time.
# max_abs_diff is read as text: as a number, kable would round 4.885e-15 to 0
# and the table would claim a stricter result than the check actually produced.
knitr::kable(read.csv("parity-summary.csv",
                      colClasses = c(max_abs_diff = "character")),
             format = "html", table.attr = 'class="table data-output"')
```

The difference is exactly zero for every design whose responses are rounded, and about
5e-15 for the unrounded continuous design, where R's CSV writer prints 15 significant
digits. At that precision the residual cannot be told apart from the writer's own rounding,
and it is what the check's 1e-6 tolerance is there to absorb. The stricter harness dumps 17
digits and resolves what is really there: a handful of cells apart by a few units in the last
place, from `exp()`, whose rounding IEEE-754 does not fix. There are no categorical mismatches
anywhere. The uniform stream is exact by construction, since every intermediate product stays
below $2^{53}$, and the harness pins every path built from IEEE-exact arithmetic alone to a
recorded hash in `tools/parity/golden.json`. The cases whose values pass through `exp()` or
`log()` are deliberately left unpinned, because those last bits belong to the maths library
rather than to the specification; the cross-language comparison still gates them on every run.
The parity check can be rerun on any new platform, so a design verified on a laptop can be
confirmed on an HPC node before a large run.

## Why this matters

Because the specification is the source of truth and the generator is shared, a design can be
authored once (in the app, in R or in Python) and reproduced everywhere. It can be preregistered
as a small JSON file that anyone can run, scaled from a laptop to a cluster without drift and
handed between collaborators who work in different languages.

This property turns an informal request to simulate some data into a reproducible, shareable
methodological artefact.
