---
title: "Cross-language reproducibility"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cross-language reproducibility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(pilotr)
```

## The challenge

A toolkit that spans R and Python faces a subtle obstacle. The two ecosystems use different
default random-number generators (R's Mersenne-Twister with inversion, NumPy's PCG64), so a
direct port produces different data from the same seed. That would undermine the guarantee
that a design authored once can be reproduced anywhere.

## A shared generator

pilotr sets aside the native RNGs and provides a shared generator implemented identically
in both languages.

- Uniform deviates come from L'Ecuyer's (1988) combined linear congruential generator,
  whose arithmetic stays below $2^{53}$ and is therefore exact in IEEE-754 doubles (R) and
  Python integers alike.
- Normal deviates use Wichura's (1988) Algorithm AS 241 inverse-CDF, the algorithm R's
  `qnorm()` uses, so deviates agree to full double precision.
- The remaining draws (Cholesky-correlated random effects, inverse-CDF Poisson and ordinal
  draws, Marsaglia–Tsang gamma draws for the Beta family, a Fisher–Yates item sampler for
  partial crossing) are derived from those two through a documented, identical consumption
  order.

As a result, given the same specification and seed, the R and Python implementations
produce bit-identical data sets. The specification, rather than the implementation, serves
as the source of truth.

## Determinism in R

The same specification and seed always give the same data.

```{r}
spec <- build_spec(list(
  name = "demo", seed = 2024, design_kind = "between", n_subject = 200,
  factor_name = "group", lev1 = "a", lev2 = "b",
  intercept = 100, effect = 5, family = "gaussian", resp_name = "score", sigma = 10))

isTRUE(all.equal(simulate_design(spec), simulate_design(spec)))
```

Changing the seed changes the draws, while the structure and the ground-truth parameters stay
fixed.

```{r}
spec2 <- spec; spec2$seed <- spec$seed + 1L
identical(simulate_design(spec)$score, simulate_design(spec2)$score)
```

## The same design in Python

The Python package reads the identical specification and produces the identical data. The
code below is not evaluated here. It shows the Python you would run, and its output matches
the R data to the last bit.

```python
from pilotr import simulate, load_spec

# the same design.json the R package (and the app) use
data = simulate("design.json")
```

The repository includes a parity check that simulates both designs in both languages and reports
the maximum absolute difference. That difference is 0.0 across thousands of rows, with no
categorical mismatches. The simulation core is also bit-identical across machines and R versions,
so a design verified on a laptop reproduces exactly on an HPC node.

## Why this matters

Because the specification is the source of truth and the generator is shared, a design can be:

- authored once (in the app, in R, or in Python) and reproduced everywhere;
- preregistered as a small JSON file that anyone can run;
- scaled from a laptop to a cluster without drift;
- handed between collaborators who work in different languages.

This property turns an informal request to simulate some data into a reproducible, shareable
methodological artifact.
