---
title: "Getting started with pilotr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with pilotr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(pilotr)
```

## One specification, three interfaces

**pilotr** is built around a single idea: a *portable design specification* that fully
describes a data-generating process and is consumed identically by three interfaces — a
no-code web app, this R package, and a Python package. The specification is the unit of
reproducibility: not a screenshot of a GUI, nor a language-specific script, but a complete,
executable description you can version-control, attach to a preregistration, share with a
collaborator, and run unchanged in R or Python to get **bit-identical** data.

This vignette walks through the core loop: **describe → simulate → inspect → export**. The
other vignettes go deeper into [power](power-analysis.html),
[precision/ROPE design analysis](precision-rope.html),
[cross-language reproducibility](cross-language.html), and the
[response families](response-families.html).

## Describe a design

The specification is a small list (serialised to JSON). You can write it by hand, point and
click in the web app, or build it from a flat set of inputs with `build_spec()`. Here is a
crossed psycholinguistic design: subjects respond to items in a *related* or *unrelated*
condition, with reaction time as the outcome, crossed by-subject and by-item random
intercepts and slopes, and a small priming effect on the log scale.

```{r}
spec <- build_spec(list(
  name = "priming", seed = 2024,
  design_kind = "within", include_items = TRUE,
  n_subject = 24, n_item = 20,
  factor_name = "condition", lev1 = "related", lev2 = "unrelated",
  intercept = 6.0, effect = 0.05,
  subj_int_sd = 0.12, subj_slope_sd = 0.04, subj_corr = 0.2,
  item_int_sd = 0.08, item_slope_sd = 0.02, item_corr = -0.1,
  family = "shifted_lognormal", resp_name = "RT", sigma = 0.30, shift = 200
))
```

The portable artifact is plain JSON — this is exactly what the web app downloads and what
the Python package reads:

```{r}
cat(spec_json(spec))
```

You can read a specification back from a file (or string) with `load_spec()`, so a design
authored anywhere runs everywhere.

## Simulate

`simulate_design()` turns a specification (a list **or** a path to a JSON file) into an
analysis-ready data frame:

```{r}
data <- simulate_design(spec)
nrow(data)
head(data)
```

The design is crossed, so every subject sees every item in both conditions
(`r 24 * 20 * 2` rows here).

## Reproducible by construction

pilotr ships a shared random-number generator, so the *same specification and seed* produce
the *same data every time* — in this R session, in a fresh one, in Python, and on any
machine:

```{r}
isTRUE(all.equal(simulate_design(spec), simulate_design(spec)))
```

Change the seed and the data change; change nothing and they never do. See the
[cross-language vignette](cross-language.html) for how this holds across R and Python.

## Export a self-contained script

`generate_r_script()` emits a stand-alone R script that embeds the specification and
reproduces the design with no external files — handy for an appendix or a preregistration:

```{r}
cat(generate_r_script(spec))
```

## Where to go next

- **[Power & design analysis](power-analysis.html)** — simulation-based power with Type S and
  Type M errors, for two-group and crossed mixed-effects designs.
- **[Precision / ROPE design analysis](precision-rope.html)** — plan by precision against a
  region of practical equivalence, and sweep sample size.
- **[Cross-language reproducibility](cross-language.html)** — the shared generator that makes
  R and Python agree to the last bit.
- **[Response families](response-families.html)** — Gaussian, reaction-time, accuracy, count,
  ordinal, and proportion outcomes.
