---
title: "Getting started with pilotr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with pilotr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(pilotr)
```

## One specification, three interfaces

pilotr is built around a portable design specification that fully describes a data-generating
process and is consumed identically by three interfaces, namely a no-code web application,
this R package and a Python package. The specification is the unit of reproducibility. It is
a complete, executable description rather than a screenshot of a graphical interface or a
language-specific script. You can place it under version control, attach it to a
preregistration, share it with a collaborator, and run it unchanged in R or Python to obtain
bit-identical data.

This vignette walks through the core loop of describe, simulate, inspect and export. The
other vignettes go deeper into [power](power-analysis.html),
[precision/ROPE design analysis](precision-rope.html),
[cross-language reproducibility](cross-language.html), and the
[response families](response-families.html).

## Describe a design

The specification is a small list serialised to JSON. You can write it by hand, point and
click in the web application, or build it from a flat set of inputs with `build_spec()`. The
example below is a crossed psycholinguistic design. Subjects respond to items in a related or
unrelated condition, with reaction time as the outcome, crossed by-subject and by-item random
intercepts and slopes, and a small priming effect on the log scale.

```{r}
spec <- build_spec(list(
  name = "priming", seed = 2024,
  design_kind = "within", include_items = TRUE,
  n_subject = 24, n_item = 20,
  factor_name = "condition", lev1 = "related", lev2 = "unrelated",
  intercept = 6.0, effect = 0.05,
  subj_int_sd = 0.12, subj_slope_sd = 0.04, subj_corr = 0.2,
  item_int_sd = 0.08, item_slope_sd = 0.02, item_corr = -0.1,
  family = "shifted_lognormal", resp_name = "RT", sigma = 0.30, shift = 200
))
```

The portable artefact is plain JSON. This is exactly what the web application downloads and
what the Python package reads.

```{r, comment = ""}
cat(spec_json(spec))
```

You can read a specification back from a file (or string) with `load_spec()`, so a design
authored anywhere runs everywhere.

## Simulate

`simulate_design()` turns a specification, supplied either as a list or as a path to a JSON
file, into an analysis-ready data frame.

```{r}
data <- simulate_design(spec)
nrow(data)
head(data)
```

The design is crossed, so every subject sees every item in both conditions
(`r 24 * 20 * 2` rows here).

## Reproducible by construction

pilotr ships a shared random-number generator, so the same specification and seed produce the
same data every time. This holds in the current R session, in a fresh one, in Python, and on
any machine.

```{r}
isTRUE(all.equal(simulate_design(spec), simulate_design(spec)))
```

Changing the seed changes the data, and leaving the specification unchanged leaves the data
unchanged. The [cross-language vignette](cross-language.html) describes how this holds across
R and Python.

## Export a self-contained script

`generate_r_script()` emits a stand-alone R script that embeds the specification and
reproduces the design with no external files. This is convenient for an appendix or a
preregistration.

```{r, comment = ""}
cat(generate_r_script(spec))
```

## Where to go next

- [Power and design analysis](power-analysis.html), covering simulation-based power with Type
  S and Type M errors, for two-group and crossed mixed-effects designs.
- [Precision / ROPE design analysis](precision-rope.html), which plans by precision against a
  region of practical equivalence and sweeps sample size.
- [Cross-language reproducibility](cross-language.html), describing the shared generator that
  makes R and Python agree to the last bit.
- [Response families](response-families.html), covering Gaussian, reaction-time, accuracy,
  count, ordinal and proportion outcomes.
