---
title: "Power and design analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Power and design analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "")
# Console colour carries no meaning on a rendered page. pkgdown turns it on for
# its own build, and the escape sequences then reach the reader as literal text,
# so colour is switched off here for a plain vignette render and a site build
# alike. The fixed width keeps printed output inside the documentation column.
options(cli.num_colors = 1, cli.hyperlink = FALSE, crayon.enabled = FALSE,
        width = 80)
```

```{r setup}
library(pilotr)
```

Simulation-based power addresses how often an analysis would detect an effect if the world
matched the specification exactly. pilotr estimates this by repeatedly simulating from the
ground-truth specification, fitting the analysis model and recording the proportion of
significant results. Beyond power, it reports the design-analysis quantities of Gelman and
Carlin (2014), namely the Type S (sign) error and the Type M (magnitude, or exaggeration)
ratio.

> The mixed-effects examples below use small `n_sims`, and their results were precomputed with
> exactly the code shown and shipped with the package, so that the vignette builds quickly.
> For real planning, we recommend `n_sims >= 200`, with more replicates for stable Type S and
> Type M estimates.

## Two-group Gaussian

The classic case has a closed-form analytic power, which the simulation matches.

```{r}
spec <- build_spec(list(
  name = "two_group", seed = 1, design_kind = "between", n_subject = 64,
  factor_name = "group", lev1 = "control", lev2 = "treatment",
  intercept = 100, effect = 5, family = "gaussian",
  resp_name = "score", sigma = 10))

pw <- power_design(spec, n_sims = 500)
unlist(pw[c("power", "type_s", "type_m", "true_effect", "mean_estimate")])
```

At roughly 50% power, the Type M ratio is well above 1. Conditional on significance, the
estimated effect is exaggerated, even though the average estimate over all replicates is
unbiased. This reflects the statistical-significance filter, which is precisely what design
analysis is meant to expose.

## Crossed mixed-effects designs

A feature that distinguishes pilotr from a marginal simulator is power estimation for crossed
by-subject and by-item designs. The R interface fits the maximal model
`y ~ cond + (1 + cond | subject) + (1 + cond | item)` with `lme4`/`lmerTest` and tests the
fixed effect with Satterthwaite degrees of freedom.

```{r}
spec_c <- build_spec(list(
  name = "priming", seed = 1, design_kind = "within", include_items = TRUE,
  n_subject = 24, n_item = 18,
  factor_name = "condition", lev1 = "related", lev2 = "unrelated",
  intercept = 6, effect = 0.06,
  subj_int_sd = 0.12, subj_slope_sd = 0.04, subj_corr = 0.2,
  item_int_sd = 0.08, item_slope_sd = 0.02, item_corr = -0.1,
  family = "shifted_lognormal", resp_name = "RT", sigma = 0.3, shift = 200))
```

The priming design above is fully crossed, so each of the 24 participants sees all 18 items in
both conditions, 864 trials in all. A counterbalanced study, in which each participant sees
each item in one condition only, has half the observations and lower power. Plan such a study
with the two-list encoding among the
[worked encodings](https://github.com/pablobernabeu/pilotr/blob/main/spec/SPEC.md#worked-encodings)
of the specification.

```{r, eval = FALSE}
# A tiny replicate count keeps the vignette fast. Use 200 or more for real planning.
pm <- power_mixed(spec_c, n_sims = 20)
unlist(pm[c("power", "type_s", "type_m", "n_converged")])
```

```{r, include = FALSE}
# The result of running exactly the chunk above, precomputed and shipped with
# the package so that the vignette builds within CRAN's check-time budget.
pm <- as.list(read.csv("power-mixed-cache.csv"))
```

```{r, echo = FALSE}
unlist(pm[c("power", "type_s", "type_m", "n_converged")])
```

`n_converged` reports how many replicates the maximal model actually fit. This is a useful
diagnostic in its own right, since convergence problems are common in small crossed designs,
and it is the denominator of `power`, which is the significant proportion among the converged
replicates.

`power_mixed()` carries its own simulation loop over the portable specification, with no other
power package underneath it. It covers territory pioneered by
[simr](https://doi.org/10.1111/2041-210x.12504) (Green and MacLeod, 2016) and
[mixedpower](https://doi.org/10.3758/s13428-021-01546-0) (Kumle, Vo and Draschkow, 2021). pilotr
differs in being driven by the cross-language specification, in reporting Type S and Type M
errors alongside power and in built-in parallelisation.

## Which model each family gets

`power_mixed()` and `precision_design()` choose the fitter by the response family, so that the
estimates are on the scale the specification's coefficients are written on. The mean estimate,
Type M and the region of practical equivalence then compare like with like. A model without
random effects, such as that of a between-subjects design with one row per subject, is fitted by
`lm()` or `glm()`. Each result names its fitter and test in `fitter`.

| Family | R `power_mixed()` | R `precision_design()` | Python `power_mixed` |
|---|---|---|---|
| `gaussian`, `exgaussian` | `lmerTest::lmer()`, Satterthwaite *t* | `lme4::lmer()`, Wald interval | MixedLM, Wald *z* |
| `lognormal`, `shifted_lognormal` | as above, on the log scale | as above, on the log scale | MixedLM on the log scale, Wald *z* |
| `bernoulli` | `lme4::glmer()`, binomial, Wald *z* | `lme4::glmer()`, Wald interval on the logit scale | linear model on the response scale, a warning, no mean estimate or Type M |
| `poisson` | `lme4::glmer()`, Poisson, Wald *z* | `lme4::glmer()`, Wald interval on the log scale | as for `bernoulli` |
| `ordinal`, `beta` | linear model on the response scale, a warning, no mean estimate or Type M, and no Type S for an interaction | linear model on the response scale, a warning, no ROPE decisions (the interval width stays) | as for `bernoulli` |

A Wald *z* treats the variance components as known. It is the test
[Bolker et al. (2009)](https://doi.org/10.1016/j.tree.2008.10.008) recommend for the fixed
effects of a GLMM without overdispersion. With few subjects or items, however, it rejects more
often than its nominal level ([Li and Redden, 2015](https://doi.org/10.1186/s12874-015-0026-x)),
so the power of a small accuracy or count design is somewhat overstated.

Ordinal and Beta outcomes have no frequentist model here yet, so their power is that of the
linear model's test and their Type S compares signs alone. A bounded response compresses
differences near the ends of its range, so an interaction can change sign between the two
scales, and its Type S is withheld. `generate_design_analysis()` writes a Bayesian design
analysis for them, on the logit scale of their coefficients. Python's `power_mixed` takes one
within factor crossed with items and fits the same model whatever else the specification
declares, as its own guide describes.

## A power curve

The following sweep over the number of subjects reports power at each size.

```{r, eval = FALSE}
curve <- power_curve_mixed(
  spec_c,
  subject_ns = c(8, 12, 16, 24, 32, 44, 56),
  n_sims = 50)
curve
```

```{r, include = FALSE}
# The result of running exactly the chunk above, precomputed and shipped with
# the package so that the vignette builds within CRAN's check-time budget.
curve <- read.csv("power-curve-cache.csv")
```

```{r, echo = FALSE}
curve
```

## The sample size the curve implies

The curve is a means to an end. What the analysis is run for is the sample size at which
power reaches the target, and that is the number a preregistration quotes. Reading it off the
table or the plot judges points whose Monte Carlo intervals overlap, and yields a bare figure
with no interval attached to it. `target_n()` fits the curve and inverts the fit, so the
crossing is estimated, and arrives with the uncertainty a simulated curve carries.

```{r}
solved <- target_n(curve, target = 0.8)
unlist(solved[c("n", "n_lo", "n_hi")])
```

Fifty replicates per point is few, and the interval says so. The fit is a binomial regression
of power on the square root of the sample size, weighted by the replicates behind each point,
and the interval is the delta-method interval that `MASS::dose.p()` computes for a fitted
`glm`. Nothing is extrapolated: a curve that never reaches the target within the sizes swept is
refused, and the refusal reports the range the sweep did cover.

For a curve swept over something other than sample size, `solve_curve()` is the general form,
with `transform = "identity"` where the axis is an effect size.

Plotting the curve and the solved size together shows what the solve has done.

```{r, fig.width = 6, fig.height = 3.6, dev.args = list(bg = "transparent")}
# The transparent device canvas is what lets the page colour through. A ggplot
# theme alone cannot do it, since the device paints white underneath. The
# website's dark mode then inverts the figure's ink, so the axes and labels
# follow the theme and the figure carries no opaque matte.
library(ggplot2)
# Each power estimate is a proportion over the converged replicates, so it
# carries a binomial Monte Carlo standard error. The shaded band is the 95%
# interval.
curve$se <- sqrt(curve$power * (1 - curve$power) / curve$n_converged)
ggplot(curve, aes(n_subject, power)) +
  geom_hline(yintercept = 0.8, linetype = 2, colour = "grey60") +
  # The solved sample size and its interval. The dashed horizontal line marks
  # the target and the band marks where the curve reaches it.
  annotate("rect", xmin = solved$lo, xmax = solved$hi, ymin = -Inf, ymax = Inf,
           fill = "grey60", alpha = .15) +
  geom_vline(xintercept = solved$value, linetype = 2, colour = "grey60") +
  geom_ribbon(aes(ymin = pmax(0, power - 1.96 * se),
                  ymax = pmin(1, power + 1.96 * se)),
              alpha = .15, fill = "#2C6FB0") +
  geom_line(colour = "#2C6FB0", linewidth = 0.8) +
  geom_point(colour = "#2C6FB0", size = 2.6) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = expression(italic(N) ~ "subjects"), y = "Power") +
  theme_minimal(base_size = 12) +
  # theme_minimal still paints a white plot.background over the transparent
  # canvas, so both surfaces have to be cleared for the page colour to reach the
  # figure. The ink is left at its default, because the website inverts the
  # figure in dark mode, which turns the dark axis text light, whereas a fixed
  # mid-grey would be inverted into a muddy tan.
  theme(plot.background  = element_rect(fill = NA, colour = NA),
        panel.background = element_rect(fill = NA, colour = NA),
        panel.grid       = element_line(colour = "grey80"))
```

The shaded horizontal band along the curve is the Monte Carlo interval, the binomial standard
error of each power estimate over its converged replicates widened to a 95% envelope. The
vertical band is the solved sample size and its own interval, which is what the dashed target
line invites the reader to guess at.

The solved size counts the participants whose data are analysed. Simulated data are complete,
with no attrition or exclusions. A study that expects to lose a proportion *p* of its
participants therefore recruits *N* / (1 − *p*), which is 67 for 60 analysed at 10% attrition.
Losses unrelated to the outcome cost power without biasing the estimates, and recruiting the
extra participants largely restores that power. Missingness that depends on the outcomes that
would have been observed, which pilotr does not simulate, can bias the estimates as well
([Rubin, 1976](https://doi.org/10.1093/biomet/63.3.581)).

## Parallel execution

Every power and precision analysis in pilotr takes a `workers` argument that spreads the
Monte Carlo replicates across local cores. Each replicate takes its own seed from
`replicate_seeds()`, so the results are identical to a serial run whatever the worker
count, and parallelisation costs nothing in reproducibility. The mixed-model fits dominate
the running time, which makes the speed-up close to linear in the number of cores. In a
sweep the worker pool is created once and reused across all sample sizes.

```{r, eval = FALSE}
power_curve_mixed(
  spec_c, subject_ns = seq(20, 60, 10), n_sims = 500, workers = 8)
```

This design answers a serial bottleneck familiar from `simr::powerCurve()`, which this
package's maintainer previously worked around by splitting the sample-size grid across
separate jobs by hand and recombining the results afterwards
([Bernabeu, 2021](https://pablobernabeu.github.io/2021/parallelizing-simr-powercurve/)).
In pilotr the same gain takes one argument.

## A bridge to the Bayesian workflow

For a confirmatory Bayesian fit, `brms_bridge()` returns a ready-to-run `brms` model. It
provides the family, the fixed and random-effects formula, and a weakly-informative prior set,
all derived from the same specification, so that the planning model and the confirmatory model
remain consistent. The formula reads the contrast columns that `model_data()` adds, so the model
runs on `model_data(spec_c, simulate_design(spec_c))` or on collected data passed through
`model_data()`. Its priors are scaled to the response and to each of those columns, here on the
log scale on which a shifted lognormal's coefficients act. Printing the result shows the model
as code to copy into a script; assigning it instead gives the same parts as `formula`, `family`
and `priors`, silently, for a caller assembling its own.

```{r}
brms_bridge(spec_c)
```

## See also

The [precision / ROPE vignette](precision-rope.html) covers design analysis against a region of
practical equivalence, where the question is whether an effect is large enough to matter.
