---
title: "Power and design analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Power and design analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "")
# Console colour carries no meaning on a rendered page. pkgdown turns it on for
# its own build, and the escape sequences then reach the reader as literal text,
# so colour is switched off here for a plain vignette render and a site build
# alike. The fixed width keeps printed output inside the documentation column.
options(cli.num_colors = 1, cli.hyperlink = FALSE, crayon.enabled = FALSE,
        width = 80)
```

```{r setup}
library(pilotr)
```

Simulation-based power addresses how often an analysis would detect an effect if the world
matched the specification exactly. pilotr estimates this by repeatedly simulating from the
ground-truth specification, fitting the analysis model and recording the proportion of
significant results. Beyond power, it reports the design-analysis quantities of Gelman and
Carlin (2014), namely the Type S (sign) error and the Type M (magnitude, or exaggeration)
ratio.

> The mixed-effects examples below use small `n_sims`, and their results were precomputed with
> exactly the code shown and shipped with the package, so that the vignette builds quickly.
> For real planning, we recommend `n_sims >= 200`, with more replicates for stable Type S and
> Type M estimates.

## Two-group Gaussian

The classic case has a closed-form analytic power, which the simulation matches.

```{r}
spec <- build_spec(list(
  name = "two_group", seed = 1, design_kind = "between", n_subject = 64,
  factor_name = "group", lev1 = "control", lev2 = "treatment",
  intercept = 100, effect = 5, family = "gaussian",
  resp_name = "score", sigma = 10))

pw <- power_design(spec, n_sims = 500)
unlist(pw[c("power", "type_s", "type_m", "true_effect", "mean_estimate")])
```

At roughly 50% power, the Type M ratio is well above 1. Conditional on significance, the
estimated effect is exaggerated, even though the average estimate over all replicates is
unbiased. This reflects the statistical-significance filter, which is precisely what design
analysis is meant to expose.

## Crossed mixed-effects designs

A feature that distinguishes pilotr from a marginal simulator is power estimation for crossed
by-subject and by-item designs. The R interface fits the maximal model
`y ~ cond + (1 + cond | subject) + (1 + cond | item)` with `lme4`/`lmerTest` and tests the
fixed effect with Satterthwaite degrees of freedom.

```{r}
spec_c <- build_spec(list(
  name = "priming", seed = 1, design_kind = "within", include_items = TRUE,
  n_subject = 24, n_item = 18,
  factor_name = "condition", lev1 = "related", lev2 = "unrelated",
  intercept = 6, effect = 0.06,
  subj_int_sd = 0.12, subj_slope_sd = 0.04, subj_corr = 0.2,
  item_int_sd = 0.08, item_slope_sd = 0.02, item_corr = -0.1,
  family = "shifted_lognormal", resp_name = "RT", sigma = 0.3, shift = 200))
```

```{r, eval = FALSE}
# A tiny replicate count keeps the vignette fast. Use 200 or more for real planning.
pm <- power_mixed(spec_c, n_sims = 20)
unlist(pm[c("power", "type_s", "type_m", "n_converged")])
```

```{r, include = FALSE}
# The result of running exactly the chunk above, precomputed and shipped with
# the package so that the vignette builds within CRAN's check-time budget.
pm <- as.list(read.csv("power-mixed-cache.csv"))
```

```{r, echo = FALSE}
unlist(pm[c("power", "type_s", "type_m", "n_converged")])
```

`n_converged` reports how many replicates the maximal model actually fit. This is a useful
diagnostic in its own right, since convergence problems are common in small crossed designs,
and it is the denominator of `power`, which is the significant proportion among the converged
replicates.

`power_mixed()` carries its own simulation loop over the portable specification, with no other
power package underneath it. It covers territory pioneered by
[simr](https://doi.org/10.1111/2041-210x.12504) (Green and MacLeod, 2016) and
[mixedpower](https://doi.org/10.3758/s13428-021-01546-0) (Kumle, Vo and Draschkow, 2021). pilotr
differs in being driven by the cross-language specification, in reporting Type S and Type M
errors alongside power and in built-in parallelisation.

## A power curve

The following sweep over the number of subjects reports power at each size.

```{r, eval = FALSE}
curve <- power_curve_mixed(
  spec_c,
  subject_ns = c(8, 12, 16, 24, 32, 44, 56),
  n_sims = 50)
curve
```

```{r, include = FALSE}
# The result of running exactly the chunk above, precomputed and shipped with
# the package so that the vignette builds within CRAN's check-time budget.
curve <- read.csv("power-curve-cache.csv")
```

```{r, echo = FALSE}
curve
```

## The sample size the curve implies

The curve is a means to an end. What the analysis is run for is the sample size at which
power reaches the target, and that is the number a preregistration quotes. Reading it off the
table or the plot judges points whose Monte Carlo intervals overlap, and yields a bare figure
with no interval attached to it. `target_n()` fits the curve and inverts the fit, so the
crossing is estimated, and arrives with the uncertainty a simulated curve carries.

```{r}
solved <- target_n(curve, target = 0.8)
unlist(solved[c("n", "n_lo", "n_hi")])
```

Fifty replicates per point is few, and the interval says so. The fit is a binomial regression
of power on the square root of the sample size, weighted by the replicates behind each point,
and the interval is the delta-method interval that `MASS::dose.p()` computes for a fitted
`glm`. Nothing is extrapolated: a curve that never reaches the target within the sizes swept is
refused, and the refusal reports the range the sweep did cover.

For a curve swept over something other than sample size, `solve_curve()` is the general form,
with `transform = "identity"` where the axis is an effect size.

Plotting the curve and the solved size together shows what the solve has done.

```{r, fig.width = 6, fig.height = 3.6, dev.args = list(bg = "transparent")}
# The transparent device canvas is what lets the page colour through. A ggplot
# theme alone cannot do it, since the device paints white underneath. The
# website's dark mode then inverts the figure's ink, so the axes and labels
# follow the theme and the figure carries no opaque matte.
library(ggplot2)
# Each power estimate is a proportion over the converged replicates, so it
# carries a binomial Monte Carlo standard error. The shaded band is the 95%
# interval.
curve$se <- sqrt(curve$power * (1 - curve$power) / curve$n_converged)
ggplot(curve, aes(n_subject, power)) +
  geom_hline(yintercept = 0.8, linetype = 2, colour = "grey60") +
  # The solved sample size and its interval. The dashed horizontal line marks
  # the target and the band marks where the curve reaches it.
  annotate("rect", xmin = solved$lo, xmax = solved$hi, ymin = -Inf, ymax = Inf,
           fill = "grey60", alpha = .15) +
  geom_vline(xintercept = solved$value, linetype = 2, colour = "grey60") +
  geom_ribbon(aes(ymin = pmax(0, power - 1.96 * se),
                  ymax = pmin(1, power + 1.96 * se)),
              alpha = .15, fill = "#2C6FB0") +
  geom_line(colour = "#2C6FB0", linewidth = 0.8) +
  geom_point(colour = "#2C6FB0", size = 2.6) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = expression(italic(N) ~ "subjects"), y = "Power") +
  theme_minimal(base_size = 12) +
  # theme_minimal still paints a white plot.background over the transparent
  # canvas, so both surfaces have to be cleared for the page colour to reach the
  # figure. The ink is left at its default, because the website inverts the
  # figure in dark mode, which turns the dark axis text light, whereas a fixed
  # mid-grey would be inverted into a muddy tan.
  theme(plot.background  = element_rect(fill = NA, colour = NA),
        panel.background = element_rect(fill = NA, colour = NA),
        panel.grid       = element_line(colour = "grey80"))
```

The shaded horizontal band along the curve is the Monte Carlo interval, the binomial standard
error of each power estimate over its converged replicates widened to a 95% envelope. The
vertical band is the solved sample size and its own interval, which is what the dashed target
line invites the reader to guess at.

## Parallel execution

Every power and precision analysis in pilotr takes a `workers` argument that spreads the
Monte Carlo replicates across local cores. Each replicate takes its own seed from
`replicate_seeds()`, so the results are identical to a serial run whatever the worker
count, and parallelisation costs nothing in reproducibility. The mixed-model fits dominate
the running time, which makes the speed-up close to linear in the number of cores. In a
sweep the worker pool is created once and reused across all sample sizes.

```{r, eval = FALSE}
power_curve_mixed(
  spec_c, subject_ns = seq(20, 60, 10), n_sims = 500, workers = 8)
```

This design answers a serial bottleneck familiar from `simr::powerCurve()`, which this
package's maintainer previously worked around by splitting the sample-size grid across
separate jobs by hand and recombining the results afterwards
([Bernabeu, 2021](https://pablobernabeu.github.io/2021/parallelizing-simr-powercurve/)).
In pilotr the same gain takes one argument.

## A bridge to the Bayesian workflow

For a confirmatory Bayesian fit, `brms_bridge()` emits a ready-to-run `brms` model. It
provides the family, the fixed and random-effects formula, and a weakly-informative prior set,
all derived from the same specification, so that the planning model and the confirmatory model
remain consistent.

```{r}
brms_bridge(spec_c)
```

## See also

The [precision / ROPE vignette](precision-rope.html) covers design analysis against a region of
practical equivalence, where the question is whether an effect is large enough to matter.
