---
title: "Power and design analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Power and design analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(pilotr)
```

Simulation-based power addresses how often an analysis would detect an effect if the world
matched the specification exactly. pilotr estimates this by repeatedly simulating from the
ground-truth specification, fitting the analysis model, and recording the proportion of
significant results. Beyond power, it reports the design-analysis quantities of Gelman and
Carlin (2014), namely the Type S (sign) error and the Type M (magnitude, or exaggeration)
ratio.

> The mixed-effects examples below use very small `n_sims` so that the vignette builds quickly.
> For real planning, we recommend `n_sims >= 200`, with more replicates for stable Type S and
> Type M estimates.

## Two-group Gaussian

The classic case has a closed-form analytic power, which the simulation matches.

```{r}
spec <- build_spec(list(
  name = "two_group", seed = 1, design_kind = "between", n_subject = 64,
  factor_name = "group", lev1 = "control", lev2 = "treatment",
  intercept = 100, effect = 5, family = "gaussian", resp_name = "score", sigma = 10))

pw <- power_design(spec, n_sims = 500)
unlist(pw[c("power", "type_s", "type_m", "true_effect", "mean_estimate")])
```

At roughly 50% power, the Type M ratio is well above 1. Conditional on significance, the
estimated effect is exaggerated, even though the average estimate over all replicates is
unbiased. This reflects the statistical-significance filter, which is precisely what design
analysis is meant to expose.

## Crossed mixed-effects designs

A feature that distinguishes pilotr from a marginal simulator is power estimation for crossed
by-subject and by-item designs. The R interface fits the maximal model
`y ~ cond + (1 + cond | subject) + (1 + cond | item)` with `lme4`/`lmerTest` and tests the
fixed effect with Satterthwaite degrees of freedom.

```{r}
spec_c <- build_spec(list(
  name = "priming", seed = 1, design_kind = "within", include_items = TRUE,
  n_subject = 24, n_item = 18,
  factor_name = "condition", lev1 = "related", lev2 = "unrelated",
  intercept = 6, effect = 0.05,
  subj_int_sd = 0.12, subj_slope_sd = 0.04, subj_corr = 0.2,
  item_int_sd = 0.08, item_slope_sd = 0.02, item_corr = -0.1,
  family = "shifted_lognormal", resp_name = "RT", sigma = 0.3, shift = 200))

pm <- power_mixed(spec_c, n_sims = 20)   # tiny for the vignette; use >= 200 in practice
unlist(pm[c("power", "type_s", "type_m", "n_converged")])
```

`n_converged` reports how many replicates the maximal model actually fit. This is a useful
diagnostic in its own right, since convergence problems are common in small crossed designs.

## A power curve

The following sweep over the number of subjects shows where power crosses a target.

```{r, fig.width = 6, fig.height = 3.6}
curve <- power_curve_mixed(spec_c, subject_ns = c(15, 30), n_sims = 12)
curve

library(ggplot2)
ggplot(curve, aes(n_subject, power)) +
  geom_hline(yintercept = 0.8, linetype = 2, colour = "grey60") +
  geom_line(colour = "#2C6FB0", linewidth = 0.8) +
  geom_point(colour = "#2C6FB0", size = 2.6) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = expression(italic(N) ~ "subjects"), y = "Power") +
  theme_minimal(base_size = 12)
```

For a high-resolution sweep, these analyses are fully parallel and scale well on a cluster.
The repository includes a SLURM array job (`hpc/precision_array.slurm`) that runs the
precision sweep across many cores.

## A bridge to the Bayesian workflow

For a confirmatory Bayesian fit, `brms_bridge()` emits a ready-to-run `brms` model. It
provides the family, the fixed and random-effects formula, and a weakly-informative prior set,
all derived from the same specification, so that the planning model and the confirmatory model
remain consistent.

```{r}
brms_bridge(spec_c)
```

## See also

The [precision / ROPE vignette](precision-rope.html) covers design analysis when the question
concerns practical significance instead of statistical significance.
