---
title: "Power and design analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Power and design analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(pilotr)
```

Simulation-based power asks a simple question: if the world is exactly as my specification
says, how often would my analysis detect the effect? pilotr answers it by repeatedly
simulating from the ground-truth spec, fitting the analysis model, and recording the
proportion of significant results — and, beyond power, it reports the **design-analysis**
quantities of Gelman and Carlin (2014): the **Type S** (sign) error and the **Type M**
(magnitude / exaggeration) ratio.

> The mixed-effects examples below use very small `n_sims` so the vignette builds quickly. For
> real planning use `n_sims >= 200` (and more replicates for stable Type S/M).

## Two-group Gaussian

The classic case has a closed-form analytic power, which the simulation matches:

```{r}
spec <- build_spec(list(
  name = "two_group", seed = 1, design_kind = "between", n_subject = 64,
  factor_name = "group", lev1 = "control", lev2 = "treatment",
  intercept = 100, effect = 5, family = "gaussian", resp_name = "score", sigma = 10))

pw <- power_design(spec, n_sims = 500)
unlist(pw[c("power", "type_s", "type_m", "true_effect", "mean_estimate")])
```

At roughly 50% power, notice the **Type M** ratio well above 1: conditional on significance,
the estimated effect is *exaggerated*, even though the average estimate over all replicates is
unbiased. This is the statistical-significance filter at work, and it is exactly what
design analysis is meant to expose.

## Crossed mixed-effects designs

The capability that distinguishes pilotr from a marginal simulator is power for **crossed**
by-subject and by-item designs. The R interface fits the maximal model
`y ~ cond + (1 + cond | subject) + (1 + cond | item)` with `lme4`/`lmerTest` and tests the
fixed effect with Satterthwaite degrees of freedom.

```{r}
spec_c <- build_spec(list(
  name = "priming", seed = 1, design_kind = "within", include_items = TRUE,
  n_subject = 24, n_item = 18,
  factor_name = "condition", lev1 = "related", lev2 = "unrelated",
  intercept = 6, effect = 0.05,
  subj_int_sd = 0.12, subj_slope_sd = 0.04, subj_corr = 0.2,
  item_int_sd = 0.08, item_slope_sd = 0.02, item_corr = -0.1,
  family = "shifted_lognormal", resp_name = "RT", sigma = 0.3, shift = 200))

pm <- power_mixed(spec_c, n_sims = 20)   # tiny for the vignette; use >= 200 in practice
unlist(pm[c("power", "type_s", "type_m", "n_converged")])
```

`n_converged` reports how many replicates the maximal model actually fit — a useful diagnostic
in its own right, since convergence problems are common in small crossed designs.

## A power curve

Sweep the number of subjects to see where power crosses a target:

```{r, fig.width = 6, fig.height = 3.6}
curve <- power_curve_mixed(spec_c, subject_ns = c(15, 30), n_sims = 12)
curve

plot(curve$n_subject, curve$power, type = "b", pch = 19, ylim = c(0, 1),
     xlab = "N subjects", ylab = "Power", col = "#2C6FB0", lwd = 2)
abline(h = 0.8, lty = 2, col = "grey")
```

For a high-resolution sweep, these analyses are embarrassingly parallel and scale well on a
cluster (the repository ships a SLURM array job that runs 1,600 maximal fits across 64 cores
in under a minute).

## A bridge to the Bayesian workflow

For a confirmatory Bayesian fit, `brms_bridge()` emits a ready-to-run `brms` model — the
family, the fixed + random-effects formula, and a weakly-informative prior set — derived from
the same specification, so the planning model and the confirmatory model never diverge:

```{r}
brms_bridge(spec_c)
```

## See also

The [precision / ROPE vignette](precision-rope.html) covers design analysis when the question
is *practical* rather than statistical significance.
