---
title: "Precision and ROPE design analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Precision and ROPE design analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(pilotr)
```

When sample sizes are large, or when the question of interest is whether an effect is
*practically* meaningful, power against a point null is often the wrong target. pilotr
implements a precision-based design analysis against a region of practical equivalence (ROPE).
This is a fast, frequentist analogue of the Bayesian approach that compares a highest-density
interval with a ROPE.

> As in the power vignette, `n_sims` is kept small here so the vignette builds quickly. For real
> planning, a value of `n_sims >= 200` is recommended.

## The idea

Over Monte Carlo replicates, and for each focal fixed effect, pilotr records the 95% confidence
interval and whether it falls determinately outside the ROPE (the effect is practically
meaningful) or entirely inside it (practical equivalence to zero), together with the expected
interval width. The interval is a Wald approximation, the estimate plus or minus 1.96 standard
errors, chosen for speed and for comparability across replicates. Sweeping sample size then locates the minimum *N* at which a focal effect reaches
a determinate decision with a target probability.

## A worked design

We reuse the crossed priming design, which represents a small priming effect on log reaction
time.

```{r}
spec_c <- build_spec(list(
  name = "priming", seed = 1, design_kind = "within", include_items = TRUE,
  n_subject = 24, n_item = 18,
  factor_name = "condition", lev1 = "related", lev2 = "unrelated",
  intercept = 6, effect = 0.05,
  subj_int_sd = 0.12, subj_slope_sd = 0.04, subj_corr = 0.2,
  item_int_sd = 0.08, item_slope_sd = 0.02, item_corr = -0.1,
  family = "shifted_lognormal", resp_name = "RT", sigma = 0.3, shift = 200))
```

So that the analysis can run from the specification alone, pilotr auto-derives the analysis
model, including the contrast columns, the response transform and the mixed-model formula.

```{r}
model_formula(spec_c)
```

## Precision at a fixed *N*

We declare the focal effect (its coefficient name and true value) and a ROPE half-width. Here
the effect lives on the log scale, and we treat anything smaller than 0.02 as practically
equivalent to zero.

```{r}
pr <- precision_design(spec_c, focal = c(effect = 0.05), rope = 0.02, n_sims = 25)
pr
```

The columns are interpreted as follows. `p_meaningful` is the probability the 95% CI lands
entirely outside the ROPE, a determinate 'meaningful' decision. `p_equivalent` is the
probability it lands entirely inside, a determinate 'equivalent' decision. `mean_ci_width` is
the expected precision.

## Sweeping sample size

```{r}
prc <- precision_curve(spec_c, focal = c(effect = 0.05), subject_ns = c(15, 30), n_sims = 15)
prc[, c("n_subject", "param", "p_meaningful", "p_equivalent", "mean_ci_width")]
```

As *N* grows the interval tightens, `p_meaningful` rises and the design reaches a determinate
decision more reliably. This curve can be used to choose the smallest *N* that meets a target, for
example '`p_meaningful` ≥ 0.90', which provides a more informative criterion than power against
a point null.

## See also

The [power vignette](power-analysis.html) covers simulation-based power and Type S/M errors. The
[getting-started vignette](getting-started.html) covers the core simulate-and-inspect loop.
