---
title: "Response families"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Response families}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 3.4)
```

```{r setup}
library(pilotr)
```

Behavioural outcomes are seldom Gaussian. pilotr maps the linear predictor
$\eta = \beta_0 + \sum_k \beta_k x_k + (\text{random effects})$ to an outcome through one of
six response families. An important consequence is that the fixed intercept and effect live on
the family's own scale. That scale is the identity for Gaussian, the log scale for reaction
times and counts, and the logit scale for accuracy, ordinal and proportion outcomes. This
vignette presents each family with scale-appropriate parameters.

The helper below builds a two-group between-subjects design and reports the group means.

```{r}
demo <- function(family, intercept, effect, n = 4000, ...) {
  spec <- build_spec(list(
    name = family, seed = 1, design_kind = "between", n_subject = n,
    factor_name = "group", lev1 = "control", lev2 = "treatment",
    intercept = intercept, effect = effect, family = family, resp_name = "", ...))
  d <- simulate_design(spec)
  y <- d[[spec$response$name]]
  list(spec = spec, data = d, y = y, by_group = tapply(y, d$group, mean))
}

library(ggplot2)
fam_hist <- function(y, fill, title, xlab) {
  ggplot(data.frame(y = y), aes(y)) +
    geom_histogram(bins = 40, fill = fill, colour = "white") +
    labs(title = title, x = xlab, y = "count") +
    theme_minimal(base_size = 12)
}
```

## Gaussian

This is the default, with $y = \eta + \varepsilon$ and residual standard deviation `sigma`. It
suits continuous, roughly symmetric outcomes such as ratings averaged over many trials or
standardised scores.

```{r}
g <- demo("gaussian", intercept = 100, effect = 5, sigma = 10)
round(g$by_group, 2)
fam_hist(g$y, "#2C6FB0", "Gaussian", "score")
```

## Shifted lognormal (reaction times)

Reaction times are right-skewed and bounded below. pilotr models them as
$y = \text{shift} + \exp(\eta + \varepsilon)$, with the effect on the log scale. An
intercept of 6 implies a typical RT near `exp(6)` ms above the shift.

```{r}
rt <- demo("shifted_lognormal", intercept = 6, effect = 0.1, sigma = 0.3, shift = 200)
round(rt$by_group, 1)
fam_hist(rt$y, "#B0402C", "Shifted lognormal (RT)", "RT (ms)")
```

## Bernoulli (accuracy)

Binary accuracy is modelled through a logit link. The intercept is the log-odds of a correct
response, and the effect is a log-odds difference between conditions.

```{r}
acc <- demo("bernoulli", intercept = 0, effect = 0.5)
round(acc$by_group, 3)   # P(correct) by group
```

## Poisson (counts)

Counts via a log link (e.g. number of fixations, errors, or events). An intercept of 1.5
implies a base rate near `exp(1.5)`.

```{r}
cts <- demo("poisson", intercept = 1.5, effect = 0.3)
round(cts$by_group, 2)   # mean count by group
table(cts$y)[1:8]
```

## Ordinal (Likert)

Ordered categorical responses via a cumulative-logit model with user thresholds. The effect
shifts the latent distribution across the thresholds.

```{r}
ord <- build_spec(list(
  name = "likert", seed = 1, design_kind = "between", n_subject = 4000,
  factor_name = "group", lev1 = "control", lev2 = "treatment",
  intercept = 0, effect = 0.8, family = "ordinal", resp_name = "rating",
  thresholds = "-2, -0.6, 0.6, 2"))
r <- simulate_design(ord)
round(prop.table(table(r$group, r$rating), 1), 2)   # category proportions by group
```

## Beta (proportions)

Bounded proportions in (0, 1) are modelled through a mean–precision parameterisation. The mean
is `logit⁻¹(η)` and `phi` is the precision, with larger values giving a tighter distribution.

```{r}
bt <- demo("beta", intercept = 0, effect = 0.8, phi = 8)
round(bt$by_group, 3)   # mean proportion by group
fam_hist(bt$y, "#2E8B57", "Beta", "proportion")
```

## Choosing a family

| Outcome | Family | Scale of the effect |
|---|---|---|
| Continuous, symmetric | `gaussian` | identity |
| Reaction time | `shifted_lognormal` | log |
| Accuracy (0/1) | `bernoulli` | logit |
| Counts | `poisson` | log |
| Likert / ordered categories | `ordinal` | logit (cumulative) |
| Proportions in (0, 1) | `beta` | logit (mean) |

These match the families that researchers fit in `lme4`, `glmmTMB` and `brms`, so a design
simulated here corresponds to the model that will later be fit. The point-and-click application
exposes all six families. The full engine, including continuous predictors, interactions,
nesting and partial crossing, is available by writing the specification directly.
