---
title: "Response families"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Response families}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# The graphics device canvas, not the ggplot theme, is what makes a figure's background
# opaque, so the transparent device is what lets the page colour show through. On the
# package website that matters twice over: the light page is warm off-white rather than
# pure white, and the dark page inverts the figure, where an opaque matte would read as a
# slab in either mode.
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 3.4,
                      dev.args = list(bg = "transparent"))
```

```{r setup}
library(pilotr)
```

Behavioural outcomes are seldom Gaussian. pilotr maps the linear predictor
$\eta = \beta_0 + \sum_k \beta_k x_k + (\text{random effects})$ to an outcome through one of
seven response families. An important consequence is that the fixed intercept and effect live on
the family's own scale. That scale is the identity for Gaussian, the log scale for reaction
times and counts, and the logit scale for accuracy, ordinal and proportion outcomes. This
vignette presents each family with scale-appropriate parameters.

The helper below builds a two-group between-subjects design and reports the group means.

```{r}
demo <- function(family, intercept, effect, n = 4000, ...) {
  spec <- build_spec(list(
    name = family, seed = 1, design_kind = "between", n_subject = n,
    factor_name = "group", lev1 = "control", lev2 = "treatment",
    intercept = intercept, effect = effect, family = family, resp_name = "", ...))
  d <- simulate_design(spec)
  y <- d[[spec$response$name]]
  list(spec = spec, data = d, y = y, by_group = tapply(y, d$group, mean))
}

library(ggplot2)
fam_hist <- function(y, fill, title, xlab) {
  ggplot(data.frame(y = y), aes(y)) +
    geom_histogram(bins = 40, fill = fill, colour = NA) +
    labs(title = title, x = xlab, y = "count") +
    theme_minimal(base_size = 12) +
    # theme_minimal still paints a white plot.background over the transparent device
    # canvas, so both surfaces have to be cleared for the page colour to reach the
    # figure. The ink stays at its default: the website inverts the figure in dark mode,
    # which turns the dark axis text light of its own accord.
    theme(plot.background  = element_rect(fill = NA, colour = NA),
          panel.background = element_rect(fill = NA, colour = NA),
          panel.grid       = element_line(colour = "grey80"))
}
```

## Gaussian

This is the default, with $y = \eta + \varepsilon$ and residual standard deviation `sigma`. It
suits continuous, roughly symmetric outcomes such as ratings averaged over many trials or
standardised scores.

```{r}
g <- demo("gaussian", intercept = 100, effect = 5, sigma = 10)
round(g$by_group, 2)
fam_hist(g$y, "#2C6FB0", "Gaussian", "score")
```

## Shifted lognormal (reaction times)

Reaction times are right-skewed and bounded below. pilotr models them as
$y = \text{shift} + \exp(\eta + \varepsilon)$, with the effect on the log scale. An
intercept of 6 implies a typical RT near `exp(6)` ms above the shift.

```{r}
rt <- demo("shifted_lognormal", intercept = 6, effect = 0.1, sigma = 0.3, shift = 200)
round(rt$by_group, 1)
fam_hist(rt$y, "#B0402C", "Shifted lognormal (RT)", "RT (ms)")
```

## Lognormal (positive continuous)

The plain `lognormal` family is the shifted lognormal without the shift,
$y = \exp(\eta + \varepsilon)$, suited to positive continuous outcomes such as per-word reading
times. As with reaction times the effect is on the log scale.

```{r}
ln <- demo("lognormal", intercept = 6, effect = 0.1, sigma = 0.3)
round(ln$by_group, 1)
fam_hist(ln$y, "#7A4FB0", "Lognormal", "reading time (ms)")
```

## Bernoulli (accuracy)

Binary accuracy is modelled through a logit link. The intercept is the log-odds of a correct
response, and the effect is a log-odds difference between conditions.

```{r}
acc <- demo("bernoulli", intercept = 0, effect = 0.5)
round(acc$by_group, 3)   # P(correct) by group
```

## Poisson (counts)

Counts via a log link (e.g. number of fixations, errors, or events). An intercept of 1.5
implies a base rate near `exp(1.5)`.

```{r}
cts <- demo("poisson", intercept = 1.5, effect = 0.3)
round(cts$by_group, 2)   # mean count by group
table(cts$y)[1:8]
```

## Ordinal (Likert)

Ordered categorical responses via a cumulative-logit model with user thresholds. The effect
shifts the latent distribution across the thresholds.

```{r}
ord <- build_spec(list(
  name = "likert", seed = 1, design_kind = "between", n_subject = 4000,
  factor_name = "group", lev1 = "control", lev2 = "treatment",
  intercept = 0, effect = 0.8, family = "ordinal", resp_name = "rating",
  thresholds = "-2, -0.6, 0.6, 2"))
r <- simulate_design(ord)
round(prop.table(table(r$group, r$rating), 1), 2)   # category proportions by group
```

## Beta (proportions)

Bounded proportions in (0, 1) are modelled through a mean–precision parameterisation. The mean
is `logit⁻¹(η)` and `phi` is the precision, with larger values giving a tighter distribution.

```{r}
bt <- demo("beta", intercept = 0, effect = 0.8, phi = 8)
round(bt$by_group, 3)   # mean proportion by group
fam_hist(bt$y, "#2E8B57", "Beta", "proportion")
```

## Choosing a family

| Outcome | Family | Scale of the effect |
|---|---|---|
| Continuous, symmetric | `gaussian` | identity |
| Reaction time | `shifted_lognormal` | log |
| Positive continuous (e.g. reading time) | `lognormal` | log |
| Accuracy (0/1) | `bernoulli` | logit |
| Counts | `poisson` | log |
| Likert / ordered categories | `ordinal` | logit (cumulative) |
| Proportions in (0, 1) | `beta` | logit (mean) |

These match the families that researchers fit in `lme4`, `glmmTMB` and `brms`, so a design
simulated here corresponds to the model that will later be fit. The point-and-click application
exposes six of the seven families. The full engine, including the plain lognormal, continuous
predictors, interactions, nesting and partial crossing, is available by writing the
specification directly, as shown next.

## Writing the specification directly

The specification is a plain list, so richer designs than the builder covers can be assembled
by hand. Starting from a `build_spec()` result, the example below adds a continuous item-level
predictor, an interaction with the categorical effect, a per-subject item subset (partial
crossing) and an extra grouping factor that nests subjects.

```{r}
spec <- build_spec(list(
  name = "reading", seed = 1, design_kind = "within", include_items = TRUE,
  n_subject = 12, n_item = 24,
  factor_name = "condition", lev1 = "related", lev2 = "unrelated",
  intercept = 6, effect = 0.05,
  subj_int_sd = 0.12, subj_slope_sd = 0.04, subj_corr = 0.2,
  item_int_sd = 0.08, item_slope_sd = 0.02, item_corr = -0.1,
  family = "lognormal", resp_name = "RT", sigma = 0.25))

spec$predictors <- list(list(name = "freq", varies_by = "item", mean = 0, sd = 1))
spec$fixed$coefficients[["effect:freq"]] <- 0.02   # an interaction with the effect
spec$units$item$per_subject <- 10                  # each subject sees 10 of the 24 items
spec$random$class <- list(over = "subject", n = 6, intercept_sd = 0.05)  # subjects nested in classes

head(simulate_design(spec))
```

The derived columns follow from the specification: `freq` is the continuous predictor and
`class` is the nesting factor. The auto-derived formula picks up the interaction as
`effect_freq` and the extra grouping factor as `(1 | class)`.

```{r}
model_formula(spec)
```

Ready-to-run specifications of this kind, including a continuous-predictor reading-time design,
a nested-clusters design and a partial-crossing design, ship in the repository's
[`spec/examples/`](https://github.com/pablobernabeu/pilotr/tree/main/spec/examples) directory,
and `load_spec()` reads any of them back.
