---
title: "The no-code app"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The no-code app}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

pilotr ships a point-and-click application over the same design specification that the R and
Python packages consume. It is a thin client: every control writes into the portable JSON
specification, which you can download and run unchanged in either package to obtain identical
data.

## Two ways to run it

Installed, on your own machine. With the package installed, launch the app locally:

```{r}
pilotr::run_app()
```

Each user runs their own R process, so a heavy power run never blocks anyone else, and
simulations use all of your local cores.

In the browser, with nothing to install. A serverless build runs entirely in the browser
through WebAssembly, with no server and nothing uploaded. Try it at the
[no-code app](https://pablobernabeu.github.io/pilotr/app/). It covers the light path: build a
design, simulate, inspect the data, estimate two-group Gaussian power and export.

## What the app does

Describe the design in the left panel: the sample sizes, the factor and its two levels, the
fixed intercept and effect, a response family, and for crossed designs the random-effect
standard deviations. Then read the tabs.

- **Design spec**: the exact JSON specification, the portable source of truth.
- **Data**: the simulated data set.
- **Summary & plot**: group summaries and a plot.
- **Power**: simulation-based two-group Gaussian power with the Type S and Type M errors,
  plus a power curve over sample size.
- **R script**: a self-contained, reproducible R script. The installed app can also verify it
  by running it in a clean R session and confirming bit-for-bit reproduction.

Changing the response family resets the intercept and effect to sensible values for that
family's scale, so a point-and-click design stays valid. The advanced 'paste a JSON spec' box
accepts designs beyond the point-and-click controls, such as continuous predictors,
interactions and nesting.

## One specification, three interfaces

The app does not own the design; the specification does. Download the specification (`.json`)
or the data (`.csv`), and run the specification unchanged in R with `simulate_design()` or in
Python with `pilotr.simulate()` for identical data. Crossed mixed-effects power (via `lme4`)
and precision/ROPE design analysis run in the installed packages rather than in the browser.
Whichever interface you use, the specification you build in the app drives all three.
