---
title: "The no-code app"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The no-code app}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "", eval = FALSE)
# Console colour carries no meaning on a rendered page. pkgdown turns it on for
# its own build, and the escape sequences then reach the reader as literal text,
# so colour is switched off here for a plain vignette render and a site build
# alike. The fixed width keeps printed output inside the documentation column.
options(cli.num_colors = 1, cli.hyperlink = FALSE, crayon.enabled = FALSE,
        width = 80)
```

pilotr ships a point-and-click application over the same design specification that the R and
Python packages consume. It is a thin client. Every control writes into the portable JSON
specification, which you can download and run unchanged in either package to obtain identical
data.

## Two ways to run it

One way is to run it on your own machine. With the package installed, launch the app locally:

```{r}
pilotr::run_app()
```

Each user runs their own R process, so a heavy power run never blocks anyone else, and
simulations use all of your local cores.

The other way needs nothing installed. A serverless build runs entirely in the browser
through WebAssembly, with no server and nothing uploaded. Try it at the
[no-code app](https://pablobernabeu.github.io/pilotr/app/). It covers the light path: build a
design, simulate, inspect the data, estimate two-group Gaussian power with the sample size its
curve solves to, and export.

## What the app does

Describe the design in the left panel: the sample sizes, the factor and its two levels, the
fixed intercept and effect, a response family, and for crossed designs the random-effect
standard deviations. Then read the tabs. 'Design spec' holds the exact JSON specification, the
portable source of truth, and 'Data' the simulated data set. 'Summary & plot' gives group
summaries and a plot. 'Power' reports simulation-based two-group Gaussian power with the Type S
and Type M errors, together with a power curve over sample size and the sample size at which
that curve reaches 0.80, with a confidence interval on it. 'R script' emits a self-contained,
reproducible R script, which the installed app can also verify by running it in a clean R
session and confirming bit-for-bit reproduction.

Changing the response family resets the intercept and effect to sensible values for that
family's scale, so a point-and-click design stays valid. The advanced 'paste a JSON spec' box
accepts designs beyond the point-and-click controls, such as continuous predictors,
interactions and nesting.

## One specification, three interfaces

The app does not own the design. The specification does. Download the specification (`.json`)
or the data (`.csv`), and run the specification unchanged in R with `simulate_design()` or in
Python with `pilotr.simulate()` for identical data. The heavier analyses stay with the installed
packages: crossed mixed-effects power in R, through `lme4` as the reference backend, and in
Python, and precision/ROPE design analysis in R. Whichever interface you use, the specification
you build in the app drives all three.
