---
title: "Worked examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Worked examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(pilotr)
```

pilotr ships one ready-to-run specification per design family. The same JSON
files drive the Python twin and the no-code app, so a design authored once runs
unchanged across all three. `pilotr_example()` lists them, and returns the path
to each for `load_spec()`.

```{r list}
pilotr_example()
```

Each specification is simulated below, showing the family it draws from and the
first rows of the data it produces. The specification format itself is covered
in the [Get started](getting-started.html) article.

```{r examples, results = "asis"}
desc <- c(
  between_2group_gaussian = "Two-group between-subjects Gaussian.",
  crossed_mixed_rt        = "Crossed by-subject and by-item reaction times, shifted lognormal.",
  beta_proportion         = "Bounded proportions through the Beta family.",
  ordinal_likert_between  = "Five-point Likert responses via a cumulative-logit model.",
  poisson_counts_between  = "Count outcomes through a log link.",
  reading_time_continuous = "A continuous predictor with a lognormal reading-time outcome.",
  nested_clusters         = "Subjects nested in higher-level clusters, an extra grouping factor.",
  partial_crossing        = "Each subject sees a sampled subset of items."
)

for (name in pilotr_example()) {
  spec <- load_spec(pilotr_example(name))
  d <- simulate_design(spec)
  blurb <- if (name %in% names(desc)) desc[[name]] else ""
  cat(sprintf("\n### %s\n\n", name))
  cat(sprintf("%s The `%s` family, %d rows.\n\n",
              blurb, spec$response$family, nrow(d)))
  cat(knitr::kable(head(d, 4)), sep = "\n")
  cat("\n\n")
}
```
