---
title: "About"
---

## Citing pilotr

If you use pilotr in published work, please cite it.

```{r citation, echo = FALSE, results = "asis"}
ver <- as.character(utils::packageVersion("pilotr"))
cat(sprintf(
"> Bernabeu, P. (2026). *pilotr: Simulate experimental and behavioural data from a portable design specification* (R and Python package version %s). https://doi.org/10.5281/zenodo.21266313\n",
  ver))
```

```{r bibtex, echo = FALSE, results = "asis"}
# Build the BibTeX entry from the installed version so it never drifts, then
# render it with a copy button and a download link. The button uses the browser
# clipboard API; the link is a self-contained data URI, so neither depends on a
# static file being shipped alongside the site.
ver <- as.character(utils::packageVersion("pilotr"))
bib <- paste(
  "@software{bernabeu2026pilotr,",
  "  author  = {Pablo Bernabeu},",
  "  title   = {{pilotr}: Simulate experimental and behavioural data from a portable design specification},",
  "  year    = {2026},",
  sprintf("  version = {%s},", ver),
  "  doi     = {10.5281/zenodo.21266313},",
  "  url     = {https://doi.org/10.5281/zenodo.21266313},",
  "}",
  sep = "\n"
)
esc <- function(x) {
  x <- gsub("&", "&amp;", x, fixed = TRUE)
  x <- gsub("<", "&lt;", x, fixed = TRUE)
  gsub(">", "&gt;", x, fixed = TRUE)
}
uri <- paste0(
  "data:application/x-bibtex;charset=utf-8,",
  utils::URLencode(bib, reserved = TRUE)
)
cat(sprintf(
'<div class="citation-bibtex">
<pre id="pilotr-bibtex"><code>%s</code></pre>
<p class="citation-bibtex-actions">
<button type="button" class="btn btn-primary btn-sm" onclick="pilotrCopyBibtex(this)">Copy BibTeX</button>
<a class="btn btn-outline-primary btn-sm" download="pilotr.bib" href="%s">Download .bib</a>
</p>
</div>
<script>
function pilotrCopyBibtex(btn) {
  var code = document.getElementById("pilotr-bibtex");
  navigator.clipboard.writeText(code.innerText).then(function () {
    var label = btn.textContent;
    btn.textContent = "Copied";
    setTimeout(function () { btn.textContent = label; }, 1500);
  });
}
</script>
', esc(bib), uri))
```

In R, `citation("pilotr")` returns the same reference.

## The developer

pilotr is developed by [Pablo Bernabeu](https://pablobernabeu.github.io), a researcher in the
Department of Education at the University of Oxford. His work spans cognitive psychology,
neuroscience, linguistics, education and research methods, with hands-on experience of
behavioural experiments, EEG, corpus analysis, computational modelling and statistics. He
develops open, reproducible research software in R and Python, and is a Fellow of the Software
Sustainability Institute. His [ORCID record](https://orcid.org/0000-0003-1083-2460) lists his
other work. pilotr and its [Python twin](https://pablobernabeu.github.io/pilotr/py/) are part
of that work, keeping a design specification reproducible and its simulated data identical
across both languages.

## Licence

pilotr is released under the [MIT licence](../LICENSE.html). The licence covers the R and
Python packages, the no-code app and the design specification alike.

## Versioning and archival

Releases are tagged on GitHub and archived on Zenodo. The concept DOI
[10.5281/zenodo.21266313](https://doi.org/10.5281/zenodo.21266313) always resolves to the
latest archived version, so a citation stays current without naming a version. The
[changelog](../news/index.html) records what changed in each release.

## Contributing and support

Bugs and feature requests are best raised on the
[GitHub issues page](https://github.com/pablobernabeu/pilotr/issues). The
[contributing guide](../CONTRIBUTING.html) describes the development setup and the
conventions the repository follows, including the cross-language random-number contract
that any change to the generative core must keep intact.
